EDB 2.0 How EAGLE improved the Epigraphic Database Bari

### 2016-02-09 source/28/EAGLE2016_Rocco_engrev.tex
rocco updated

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{EDB 2.0 
How Eagle Europeana project improved the Epigraphic Database Bari}
\author[uniba]{Anita Rocco \corref{first}}
\address[uniba]{Dip. di Scienze dell'Antichità e del Tardo Antico, Università degli Studi di Bari ``Aldo Moro''}
\cortext[first]{Corresponding author. Email: anita.rocco@uniba.it}
\begin{document}
\maketitle
\begin{abstract}
This paper is dedicated to the evolution of the Epigraphic Database Bari (EDB) from the minimalistic design of its
origins to its current status. Although EDB, the database of inscriptions by Christians from Rome, dates back to the
late 1980s , involvement in the EAGLE – Europeana project has had a significantly positive impact on its development.
In fact, maintaining its peculiar character, dictated by its own history and, mostly, by the characteristics of its
documentary base, it has taken advantage of the solutions adopted to integrate different archives and purpose-built
best practices. 
\end{abstract}
\keywords{Epigraphic database, EDB, EAGLE Europeana, Christian inscriptions, Late antique inscriptions }

@@

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img001.png} 
\caption{The record of a bilingual inscription (ICVR V, 15147) as it appeared in data processing program ICVR in MS-DOS
in 1988.}
\label{fig:1}
\end{figure}

In addition to the text, the data processing program recorded for each inscription\footnote{ The program didn’t allow
the use of Greek fonts; in order to add texts in the Greek alphabet it was necessary to type Latin equivalent letters
in the MS Word \emph{Symbol} font, as you can see in Fig.~\ref{fig:1}}: bibliographic data (Progressivo = ICVR Number), origin
(Provenienza), type of support and technique of execution (Supp. e tecn.), function (Funzione), the presence or
absence of Christograms (\emph{Signa Christi}), and dating (Datazione). All this information was expressed by alphanumeric
codes of few characters, according to a limitation imposed by the program (Fig.~\ref{fig:1}).

\begin{figure}[hbp]
\centering
 \includegraphics[scale=0.4]{EAGLE2016Roccoengrev-img002.jpg} 
\caption{The inscription ICVR V, 15147 as it appears now in EDB, with the acronyms expanded, the image and the text
displayed in Greek characters.}
\label{fig:2}
\end{figure}

Adding \emph{metadata} to the text, even if in the minimalistic form of alphanumeric codes, accomplished the goal of describing
the epigraphic object in its widest sense as an inscribed artifact. This feature, in particular relating to
geographical information, was even more meaningful according to the characteristics of a large part of the inscriptions
recorded in EDB. In fact, the original pertinence to a monument/container (catacomb) or to a particular area of it,
allows, with reasonable certainty, to determine the patronage by a member of the Christian community and moreover to
determine the chronology, even without specific references inside the text\footnote{\citet{carletti_inscriptiones_1994}; \citet[viii-ix]{felle_introduzione_1997}.}. Likewise reporting the presence of Christological monograms allows us to identify them as explicit symbols
of Christian faith and as chronological indicators\footnote{ The monogram consisting of the first two letters of the
name \begin{otherlanguage*}{greek}Χριστός\end{otherlanguage*} can be considered the archetype of these signs. They first appear in inscriptions at the
beginning of the fourth century.}.

@@

@@

To complete information on spatial data, all cemeterial contexts have been georeferenced, so that clicking on their name
opens a new window in \emph{Google Map} that shows the modern entrance to the cemetery, with the address and its coordinates
(Fig.~\ref{fig:6}).

 \begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img006.png} 
\caption{The \emph{original context} maps.}
\label{fig:6}
\end{figure}

As a case of study, for the inscriptions \emph{in situ} pertaining to the Domitilla catacomb (Via Ardeatina) - the only
cemeterial complex with a georeferenced plan of almost its entire extension\footnote{\emph{START-Projekt: Die
Domitilla-Katakombe in Rome (Institut für Kulturgeschichte der Antike - Österreichische Akademie der Wissenschaften)}.
\citet{orlandi_case_2014}.} - clicking on the alphanumeric code\footnote{ The codes are made up of a majuscule letter
relating to a region of the catacomb \ and by another element, digit or minuscule letter, relating to a precise
internal position, respectively a gallery or a \emph{cubiculum}. } associated to the precise position of the inscriptions,
gallery or \emph{cubiculum} (F04, in Fig.~\ref{fig:6}), opens the plan of the specific area (\emph{regio}) in which the inscription is found.
In every plan, the inscriptions preserved in situ are placed and marked by ICVR and EDB number (Fig.~\ref{fig:6})\footnote{The
plans are available on START-Projektwebsite: \url{www.oeaw.ac.at/antike/index.php\&id=431}, clicking on EDB number opens a
window with the EDB record. }.

@@

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img007.png}
\caption{Conservation input fields.}
\label{fig:7}
\end{figure}

Adopting the best practice suggested by the EAGLE consortium, georeferencing is guaranteed even for the Conservation
data, linking every City/Town to GeoNames site\footnote{ \url{www.geonames.org}}, which allows the user to pinpoint the
location and to avoid ambiguity between homonyms. A link to the Trismegistos Collection\footnote{
\url{www.trismegistos.org}}, a database of papyrological and epigraphic collections, if available, also helps to identify
uniquely the place of conservation, as well as to obtain additional information, included the geographic
positioning\footnote{\url{http://www.eagle-network.eu/wp-content/uploads/2013/06/EAGLE\_D2.2.2\_Content-harmonisation-guidelines-including-GIS-and-terminologies-Second-Release.pdf}}.

@@

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img011.png}
\caption{Quick and Advanced Search masks.}
\label{fig:11}
\end{figure}

@@

\nocite{felle_fenomeni_2007,felle_esperienze_2012,jory_problems_1975}

@@

\section*{Corpora}
\begin{description}
\item[ CIL ] = \emph{Corpus Inscriptionum Latinarum}, \ Berolini 1863 ss.
\item[CIG ] = \emph{Corpus Inscriptionum Grecarum}, I-IV, Berolini 1828-1877
\item[IC] = \emph{Inscriptiones christiane urbis Romae septimo saeculo antiquiores}, I-II, G.B. de Rossi (a c.), Romae 1857-1861; \emph{Supplementum}, G. Gatti (a c.), Romae 1915. 
\item[IG] = \emph{Inscriptiones Graecae}, Berolini 1873 ss.
\item[ICVR] = \emph{Inscriptiones christianae urbis Romae septimo saeculo antiquiores. Nova series}, I-X, A. Silvagni, A. Ferrua, D. Mazzoleni, C. Carletti (a c.), Romae - In Civitate Vaticana 1922 ss.
\item[IGVR] = \emph{Inscriptiones grecae urbis Romae}, I-IV, L. Moretti (a c.), Roma 1968-1990.

>>>>

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{EDB 2.0 
How EAGLE improved the Epigraphic Database Bari}
\author[uniba]{Anita Rocco \corref{first}}
\address[uniba]{Università degli Studi di Bari ``Aldo Moro''}
\cortext[first]{Corresponding author. Email: anita.rocco@uniba.it}
\begin{document}
\maketitle
\begin{abstract}
This paper is dedicated to the evolution of the Epigraphic Database Bari (EDB) from the minimalistic design of its
origins to its current status. Although EDB, the database of inscriptions by Christians from Rome, dates back to the
late 1980s , involvement in the EAGLE – Europeana project has had a significantly positive impact on its development.
In fact, maintaining its peculiar character, dictated by its own history and, mostly, by the characteristics of its
documentary base, it has taken advantage of the solutions adopted to integrate different archives and purpose-built
best practices. 
\end{abstract}
\keywords{Epigraphic database, EDB, EAGLE Europeana, Christian inscriptions, Late antique inscriptions }

@@

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{Fig1NEW.png} 
\caption{The record of a bilingual inscription (ICVR V, 15147) as it appeared in data processing program ICVR in MS-DOS
in 1988.}
\label{fig:1}
\end{figure}

In addition to the text, the data processing program recorded for each inscription\footnote{ The program didn’t allow
the use of Greek fonts; in order to add texts in the Greek alphabet it was necessary to type Latin equivalent letters
in the MS Word \emph{Symbol} font, as you can see in Fig.~\ref{fig:1}}: bibliographic data (Progressivo = ICVR Number), origin
(Provenienza), type of support and technique of execution (Supp. e tecn.), function (Funzione), the presence or
absence of Christograms (\emph{Signa Christi}), and dating (Datazione). All this information was expressed by alphanumeric
codes of few characters, according to a limitation imposed by the program (Fig.~\ref{fig:1}).
%
%\begin{figure}[hbp]
%\centering
% \includegraphics[scale=0.4]{Fig.2.png} 
%\caption{The inscription ICVR V, 15147 as it appears now in EDB, with the acronyms expanded, the image and the text
%displayed in Greek characters.}
%\label{fig:2}
%\end{figure}
%

Adding metadata to the text, even if in the minimalistic form of alphanumeric codes, accomplished the goal of describing
the epigraphic object in its widest sense as an inscribed artifact. This feature, in particular relating to
geographical information, was even more meaningful according to the characteristics of a large part of the inscriptions
recorded in EDB. In fact, the original pertinence to a monument/container (catacomb) or to a particular area of it,
allows, with reasonable certainty, to determine the patronage by a member of the Christian community and moreover to
determine the chronology, even without specific references inside the text\footnote{\citet{carletti_inscriptiones_1994}; \citet[viii-ix]{felle_introduzione_1997}.}. Likewise reporting the presence of Christological monograms allows us to identify them as explicit symbols
of Christian faith and as chronological indicators\footnote{ The monogram consisting of the first two letters of the
name \begin{otherlanguage*}{greek}Χριστός\end{otherlanguage*} can be considered the archetype of these signs. They first appear in inscriptions at the
beginning of the fourth century.}.

@@

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{Fig4.png}
\caption{Bibliographic data fields}
\label{fig:3b}
\end{figure}

@@

To complete information on spatial data, all cemeterial contexts have been georeferenced, so that clicking on their name
opens a new window in \emph{Google Map} that shows the modern entrance to the cemetery, with the address and its coordinates.

 \begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img006.png} 
\caption{The original context maps.}
\label{fig:6}
\end{figure}

As a case of study, for the inscriptions \emph{in situ} pertaining to the Domitilla catacomb (Via Ardeatina) - the only
cemeterial complex with a georeferenced plan of almost its entire extension\footnote{\emph{START-Projekt: Die
Domitilla-Katakombe in Rome (Institut für Kulturgeschichte der Antike - Österreichische Akademie der Wissenschaften)}.
\citet{orlandi_case_2014}.} - clicking on the alphanumeric code\footnote{ The codes are made up of a majuscule letter
relating to a region of the catacomb \ and by another element, digit or minuscule letter, relating to a precise
internal position, respectively a gallery or a \emph{cubiculum}. } associated to the precise position of the inscriptions,
gallery or \emph{cubiculum} (F04, in Fig.~\ref{fig:6}), opens the plan of the specific area (\emph{regio}) in which the inscription is found.
In every plan, the inscriptions preserved \emph{in situ} are placed and marked by ICVR and EDB number (Fig.~\ref{fig:6})\footnote{The
plans are available on START-Projektwebsite: \url{www.oeaw.ac.at/antike/index.php\&id=431}, clicking on EDB number opens a
window with the EDB record. }.

@@

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{Fig7.png}
\caption{Conservation input fields.}
\label{fig:7}
\end{figure}

Adopting the best practice suggested by the EAGLE consortium, georeferencing is guaranteed even for the Conservation
data, linking every City/Town to GeoNames site\footnote{ \url{www.geonames.org}}, which allows the user to pinpoint the
location and to avoid ambiguity between homonyms. A link to the Trismegistos Collection\footnote{
\url{www.trismegistos.org}}, a database of papyrological and epigraphic collections, if available, also helps to identify
uniquely the place of conservation, as well as to obtain additional information, included the geographic
positioning.\footnote{\url{http://www.eagle-network.eu/wp-content/uploads/2013/06/EAGLE\_D2.2.2\_Content-harmonisation-guidelines-including-GIS-and-terminologies-Second-Release.pdf}}

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{Fig8.png}
\caption{Conservation maps and links}
\label{fig:8new}
\end{figure}

@@

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img011.png}
\caption{Quick and Advanced Search masks.}
\label{fig:11}
\end{figure}

@@

\nocite{felle_fenomeni_2007}
\nocite{felle_esperienze_2012}
\nocite{jory_problems_1975}
\nocite{Rocco2015a}

@@

\section*{Corpora}
\begin{description}
\item[CIL] = \emph{Corpus Inscriptionum Latinarum}, \ Berolini 1863 ss.
\item[CIG] = \emph{Corpus Inscriptionum Grecarum}, I-IV, Berolini 1828-1877
\item[IC] = \emph{Inscriptiones christiane urbis Romae septimo saeculo antiquiores}, I-II, G.B. de Rossi (a c.), Romae 1857-1861; \emph{Supplementum}, G. Gatti (a c.), Romae 1915. 
\item[IG] = \emph{Inscriptiones Graecae}, Berolini 1873 ss.
\item[ICVR] = \emph{Inscriptiones christianae urbis Romae septimo saeculo antiquiores. Nova series}, I-X, A. Silvagni, A. Ferrua, D. Mazzoleni, C. Carletti (a c.), Romae - In Civitate Vaticana 1922 ss.
\item[IGVR] = \emph{Inscriptiones grecae urbis Romae}, I-IV, L. Moretti (a c.), Roma 1968-1990.
